We are not the 99 percent: quantifying asphericity in the distribution of Local Group satellites

### 2018-01-31 paper/paper.tex
edit

The normalized quantities for different systems are well reproduced by
a multivariate gaussian distribution. 
We fit the parameters of this gaussian to the results from 
LG samples coming from three different simulations: ELVIS, Illustris1
and Illustris-DarkMatter.
We finally compare the observational results against the distributions
derived from the simulations.

We find that in the best case (Illustris1-DarkMatter) the deviation from aesphericity in the observed LG is only
expected in $3\pm2$ pairs out of a sample of $10^4$ isolated pairs. 
This places the LG as a $3\sigma$ outlier. 
The weight to explain this atypical result is not distributed equally
between the MW and M31. While M31 presents a fully typical asphericity
in the expectations from LCDM, the MW shows aspherical deviations in
plane width and the major-to-minor axis ratio highly atypical in the
framework of LCDM. 
We estimate that with the M31 $XX$ out of $2000$ pairs show normalized
characteristics larget r equal than M31, while this number drops to 
$XX$ for the MW.

The method we present to quantify how atypical is the LG is robust to
the numerical simulations used to define the LCDM expectations. 
Quantifying the degree of asphericity highlights the atypicality of
the Milky Way stressing the stark difference with M31 and its more
typical nature. 
This atypical distribution also adds up to the presence of two bright
satellite galaxies, which is also uncommon both in simulations and
observations.

The extension of the framework we present in this paper to describe
outliers found as a second order deviation (i.e. fiding a \emph{subset} of
satellites in M31 that are in a plane or \emph{adding} velocity
information) could also be possible along the line of work presented
by, that avoid finding in observations the exact observed distribution
and instead focuse in defining atypicallity.

The quantitative study we present allows us to estimate  the volumes 
that cosmological simulations have to probe in order to study similar
LG configurations.
Such atypical satellite distribution should be seen as an opportunity
to constraint in high detail the initial conditions and environment
that allowed such pattern to emerge.

>>>>

The normalized quantities for different systems are well reproduced by
a multivariate gaussian distribution. 
We fit the parameters of this gaussian to the results from 
LG samples coming from three different simulations
to finally compare the observational results against the distributions
derived from the simulations.

We find that in the best case (Illustris1-DarkMatter) the deviation
from aesphericity in the observed LG is only  expected in $3\pm2$
pairs out of a sample of $10^4$ isolated pairs. This places the LG as
a $3\sigma$ outlier.   
The weight to explain this atypical result is not distributed equally
between the MW and M31. 
While M31 presents a fully typical asphericity in the expectations
from LCDM, the MW shows aspherical deviations in plane width and the
major-to-minor axis ratio highly atypical in the 
framework of LCDM. 
We estimate that with the M31 between $37\%$ and $58\%$ of the pairs
pairs show normalized characteristics larger than M31, while
this fraction drops to less than $1\%$ for the MW.
These fractions are robust to changes in the numerical simulations and to
the criteria used to define the pairs.

The highly aspherical distribution of the brightest MW satellites is
in stark contrast to the more spherical M31 distribution.
Furthermore, the atipicallity of the MW distribution in the LCDM
context adds up to the also atypical presence of two bright satellite
galaxies.

The focus of our approach is building explicit probability
distributions for the observables of interest, instead of trying to
find simulated objects that fullfil different observational criteria. 
This approach is particularly useful in the case of atypical
observables. 
Building the probability distribution allows for an atypicallity
quantification without needing to build explicit samples of objects
that are already scarce and difficult to find in simulations.

An extension of this framework to outliers in higher order deviations
(i.e. a \emph{subset} of satellites that are in a plane or
coherent \emph{velocity} strictures) should also be possible, similar
to the work presented by XX on the satellite velocity anisotropy.

From the results we derive in this paper we also estimate that a
cosmological volume of at least $( XX )^3$ is required to find a halo pair
that shows similar asphericity characteristics as the LG. 
This atypical satellite distribution should be seen as an opportunity
to constrain in high detail the initial conditions and environment
that allowed such a pattern to emerge.
